Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-09-24 paper/tidy-missing-data-paper.Rmd
more minor changes to include pkg and proglang

---
documentclass: jss
author:
  - name: Nicholas Tierney
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics
      E774, Menzies Building
      Monash University, Clayton
    email: \email{nicholas.tierney@gmail.com}
    url: http://www.njtierney.com
  - name: Dianne Cook
    affiliation: Monash University
title:
  formatted: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # If you use tex in the formatted title, also supply version without
  plain:     "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # For running headers, if needed
  short:     "Explore missingness with \\pkg{naniar}"
abstract: >
  Despite the large body of research on missing value distributions and imputation, there is comparatively little literature with a focus on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with the goal of integrating missing value handling as a key part of data analysis workflows. We define a new data structure, and a suite of new operations. Together, these provide a connected framework for handling, exploring, and imputing missing values. These methods are available in the R package \pkg{naniar}.
keywords:
  # at least one keyword must be supplied
  formatted: [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, "\\proglang{R}"]
  plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, R]
preamble: >
  \usepackage{amsmath}
  \usepackage[english]{babel}
  \usepackage{float}
  \usepackage{booktabs}
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    keep_tex: yes
    citation_package: 'natbib'
# bibliography: tidy-missing-data-paper.bib
bibliography: "tidy-missing-data-paper-no-jss.bib"
cls: jss.cls
---

@@

This paper is organized in the following way. Section (\@ref(background)) provides the background to tidy data principles and tools (\@ref(tidy-data-concepts)) and missing data representations (\@ref(missing-data-rep-dep)). Section \@ref(existing-software) summarises existing software for handling missing data. The new work extending the tidy tools to facilitate exploring, visualizing, and imputing missing data is discussed in Section \@ref(extensions). Graphics (\@ref(graphics)) and Numerical summaries (\@ref(num-sum)) of missing values are then discussed. An application illustrating the use of the new methods is shown in Section \@ref(case-study). Section \@ref(discussion) discusses strengths, limitations, and future directions.

@@

Data manipulation is made input- and output-tidy with R packages \pkg{dplyr} and \pkg{tidyr} [@dplyr; @tidyr]. These provide the five "verbs" of data manipulation: data reshaping, sorting, filtering, transforming, and aggregating. Data reshaping goes from long to wide formats; sorting arranges rows in a specific order; filtering removes rows based on a condition; transforming, changes existing variables or adds new ones; aggregating creates a single value from many values, say, for example, in computing the minimum, maximum, and mean.

@@

Modelling tools work well with tidy data, as they have a clear mapping from variables in the data to the formula for a model. For example in R, y regressed on x and z is: \code{lm(y ~ x + z)}. Modelling tools are input tidy, but their output is always messy - it is not in the right format for subsequent steps in analysis. For example, estimated coefficients, predictions, and residuals from one model cannot be easily combined with the output of another model. Messy models have been partially addressed with the \pkg{broom} package, which tidies up model outputs into a tidy data format for data analysis, and the developing \pkg{recipes} package, which helps make modelling input- and output-tidy [@recipes].

@@

\pkg{VIM} [@VIM] implements well-used imputation methods K nearest neighbors, regression, hot-deck, and iterative robust model-based imputation. These diverse approaches allows for imputing with semi-continuous, continuous, count, and categorical data. VIM identifies imputed cases by adding an indicator variable with a suffix "_imp". So "Var1" has a sibling column, "Var1_imp", with values TRUE or FALSE indicating imputation. \pkg{VIM} also has a variety of visualization methods, discussed in \@ref(exploration). \pkg{simputation} provides an interface to imputation methods from VIM, in addition providing hotdeck imputation, and the EM algorithm [@simputation; @Dempster1977]. \pkg{simputation} provides a consistent formula interface for all imputation methods, and always returns a dataframe with the updated imputations. \pkg{Hmisc} [@Hmisc], provides predictive mean matching, \pkg{imputeTS} [@imputeTS], provides time series imputation methods, and \pkg{missMDA} [@missMDA], imputes data using principal components analysis.

@@

\pkg{Amelia} [@amelia] assumes data are multivariate normal, and samples from the posterior, and allows for incorporation of information on the values in a prior. It uses the computationally efficient (and parallelizable) Expectation-Maximization Bootstrap (EMB) algorithm [@Honaker2010]. \pkg{norm} [@norm], provides multiple imputation using EM for multivariate normal data, drawing from methods in the NORM software [@schafer-norm]. \pkg{norm} does not provide a framework for tracking missing values, instead providing tools for making inference from multiple imputation.

@@

```{r gg-box-na, fig.height = 3, fig.width = 9, out.width = "100%", fig.align = "center", fig.cap = "ggplot2 provides different visualizations depending on what type of data has missing values for data of student test scores in school year. (A) Data is complete; (B) Missings are only in year - an additional 'NA' boxplot is created; (C) Missings only in scores, no additional missingness information is shown; (D) Missings in both scores and year, additional missing information is shown. The missingness category is only shown when there are missings in categorical variables such as year (plots (B) and (D)). In (C), no missingness information is given on the graphic, despite there being missings in score, and a warning message is displayed about the number of missing values omitted.", fig.show = "asis"}

@@

Applying tidyverse principles has the potential to clarify missing data exploration, visualization, and imputation. This section discusses how these principles are applied for data structures (\@ref(data-structure)), common operations (verbs) (\@ref(verbs)), graphics (\@ref(graphics)), and data summaries (\@ref(num-sum)). Care has been taken to make the names and design these features intuitive for their purpose, and is discussed throughout.

@@

<!-- since 0 and 1 could interchangably indicate the absence (0) of observation, or the presence (1) of a missing value. -->

>>>>

---
documentclass: jss
author:
  - name: Nicholas Tierney
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics
      E774, Menzies Building
      Monash University, Clayton
    email: \email{nicholas.tierney@gmail.com}
    url: http://www.njtierney.com
  - name: Dianne Cook
    affiliation: Monash University
title:
  formatted: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # If you use tex in the formatted title, also supply version without
  plain:     "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # For running headers, if needed
  short:     "Explore missingness with \\pkg{naniar}"
abstract: >
  Despite the large body of research on missing value distributions and imputation, there is comparatively little literature with a focus on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with the goal of integrating missing value handling as a key part of data analysis workflows. We define a new data structure, and a suite of new operations. Together, these provide a connected framework for handling, exploring, and imputing missing values. These methods are available in the \proglang{R} package \pkg{naniar}.
keywords:
  # at least one keyword must be supplied
  formatted: [statistical computing, statistical graphics, data science, data visualization, "\\pkg{tidyverse}", data pipeline, "\\proglang{R}"]
  plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, R]
preamble: >
  \usepackage{amsmath}
  \usepackage[english]{babel}
  \usepackage{float}
  \usepackage{booktabs}
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    keep_tex: yes
    citation_package: 'natbib'
# bibliography: tidy-missing-data-paper.bib
bibliography: "tidy-missing-data-paper-no-jss.bib"
cls: jss.cls
---

@@

This paper is organized in the following way. Section \@ref(background) provides the background to tidy data principles and tools \@ref(tidy-data-concepts) and missing data representations (\@ref(missing-data-rep-dep)). Section \@ref(existing-software) summarises existing software for handling missing data. The new work extending the tidy tools to facilitate exploring, visualizing, and imputing missing data is discussed in Section \@ref(extensions). Graphics (\@ref(graphics)) and Numerical summaries (\@ref(num-sum)) of missing values are then discussed. An application illustrating the use of the new methods is shown in Section \@ref(case-study). Section \@ref(discussion) discusses strengths, limitations, and future directions.

@@

Data manipulation is made input- and output-tidy with \proglang{R} packages \pkg{dplyr} and \pkg{tidyr} [@dplyr; @tidyr]. These provide the five "verbs" of data manipulation: data reshaping, sorting, filtering, transforming, and aggregating. Data reshaping goes from long to wide formats; sorting arranges rows in a specific order; filtering removes rows based on a condition; transforming, changes existing variables or adds new ones; aggregating creates a single value from many values, say, for example, in computing the minimum, maximum, and mean.

@@

Modelling tools work well with tidy data, as they have a clear mapping from variables in the data to the formula for a model. For example in \proglang{R}, y regressed on x and z is: \code{lm(y ~ x + z)}. Modelling tools are input tidy, but their output is always messy - it is not in the right format for subsequent steps in analysis. For example, estimated coefficients, predictions, and residuals from one model cannot be easily combined with the output of another model. Messy models have been partially addressed with the \pkg{broom} package, which tidies up model outputs into a tidy data format for data analysis, and the developing \pkg{recipes} package, which helps make modelling input- and output-tidy [@recipes].

@@

\pkg{VIM} [@VIM] implements well-used imputation methods K nearest neighbors, regression, hot-deck, and iterative robust model-based imputation. These diverse approaches allows for imputing with semi-continuous, continuous, count, and categorical data. VIM identifies imputed cases by adding an indicator variable with a suffix "\code{_imp}". So, "\code{Var1}" has a sibling column, "\code{Var1_imp}", with values TRUE or FALSE indicating imputation. \pkg{VIM} also has a variety of visualization methods, discussed in \@ref(exploration). \pkg{simputation} provides an interface to imputation methods from \pkg{VIM}, in addition providing hotdeck imputation, and the EM algorithm [@simputation; @Dempster1977]. \pkg{simputation} provides a consistent formula interface for all imputation methods, and always returns a dataframe with the updated imputations. \pkg{Hmisc} [@Hmisc], provides predictive mean matching, \pkg{imputeTS} [@imputeTS], provides time series imputation methods, and \pkg{missMDA} [@missMDA], imputes data using principal components analysis.

@@

\pkg{Amelia} [@amelia] assumes data are multivariate normal, and samples from the posterior, and allows for incorporation of information on the values in a prior. It uses the computationally efficient (and parallelizable) Expectation-Maximization Bootstrap (EMB) algorithm [@Honaker2010]. \pkg{norm} [@norm], provides multiple imputation using EM for multivariate normal data, drawing from methods in the \proglang{NORM} software [@schafer-norm]. \pkg{norm} does not provide a framework for tracking missing values, instead providing tools for making inference from multiple imputation.

@@

```{r gg-box-na, fig.height = 3, fig.width = 9, out.width = "100%", fig.align = "center", fig.cap = "\\pkg{ggplot2} provides different visualizations depending on what type of data has missing values for data of student test scores in school year. (A) Data is complete; (B) Missings are only in year - an additional 'NA' boxplot is created; (C) Missings only in scores, no additional missingness information is shown; (D) Missings in both scores and year, additional missing information is shown. The missingness category is only shown when there are missings in categorical variables such as year (plots (B) and (D)). In (C), no missingness information is given on the graphic, despite there being missings in score, and a warning message is displayed about the number of missing values omitted.", fig.show = "asis"}

@@

Applying \pkg{tidyverse} principles has the potential to clarify missing data exploration, visualization, and imputation. This section discusses how these principles are applied for data structures (\@ref(data-structure)), common operations (verbs) (\@ref(verbs)), graphics (\@ref(graphics)), and data summaries (\@ref(num-sum)). Care has been taken to make the names and design these features intuitive for their purpose, and is discussed throughout.

@@
